---
title: "Introduction to structlearn: Structure Learning by KL/IC"
output: rmarkdown::html_vignette
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Mikkel Meyer Andersen
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introduction}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width = 7, cache = FALSE)
```

## Introduction

First, the library is loaded:
```{r, message=FALSE}
library(structlearn)
```


```{r}
data(danedb)
db <- danedb
db[, "DYS389II"] <- db[, "DYS389II"] - db[, "DYS389I"]
```

```{r}
d <- prepare_data(db)
print(d)
plot(d)
```

```{r}
clfit <- chow_liu(d)
clfit
plot(clfit, layout = layout_as_tree)
```

```{r}
cl_to_tcherry(clfit)
```



```{r}
# library(bnlearn)
# db_fact <- data.frame(apply(db, 2, as.character), stringsAsFactors = TRUE)
# out <- capture.output(chow_liu_fit_undirected <- chow.liu(db_fact, mi = "mi", debug = TRUE))
# out
# 
# plot(chow_liu_fit_undirected, layout = layout_as_tree)
# 
# chow_liu_fit <- cextend(chow_liu_fit_undirected) # To get a DAG, extension is needed
# plot(chow_liu_fit, layout = layout_as_tree)
``` 

Chapter 3
On the Approximation of a Discrete
Multivariate Probability Distribution Using
the New Concept of t-Cherry Junction Tree

3.3.3  The Relation Between the Approximations Associated to the First-Order Dependence Tree and t-Cherry-Junction Tree

Cherry trees:

(Vertices, Edges, Cherries)

Cherry: (i, j, k), j middle, edge i-j-k
Only cherry with two vertices is o--o

From a cherry tree we can obtain another cherry tree by adding a new vertex, j, and
connecting the new vertex with two already existing vertices i and k by two new edges (i-j and j-k).
These two edges constitute a cherry (i, j, k). So

   j
   o
  / \
 o   o
 i   k
is a cherry tree with cherry (i, j, k).

Note that
   j
   o
  /
 o 
 i
is the only cherry tree with two vertices.

A new cherry tree can be obtained as described above.

Cherry tree not uniquely determined by its graph. 
The same graph can represent more cherry trees.

Note that in a cherry tree there is exactly one edge which is not part of a
cherry.


t-cherry tree:
Above we had

> From a cherry tree we can obtain another cherry tree by adding a new vertex, j, and
connecting the new vertex with two already existing vertices i and k by two new edges (i-j and j-k).
These two edges constitute a cherry (i, j, k).

If we instead use

> By connecting a new vertex of V by two new edges to two connected vertices
of a cherry tree, one obtains a new cherry tree.

we get a t-cherry tree instead of a cherry tree.


Remark 3.2. From a set of n vertices we obtain a cherry tree with n - 2 cherries.

Remark 3.5. A pair of adjacent vertices from the t-cherry tree may be used several
times, for connecting new vertices to them.

Algorithm 2.

```{r}
#1) the spanning tree behind the Chow–Liu
#2) construct a t-cherry tree from this spanning tree
#3) Algorithm 1: assign a t-cherry-junction tree to this t-cherry tree

# Algorithm 2.:
# 1. The first cherry of the t-cherry tree let be defined by any three vertices of the
# spanning tree which are connected by two edges.
# 2. We add a new cherry to the t-cherry tree by taking a new vertex of the spanning
# tree adjacent to the so far constructed t-cherry tree.
# 3. We repeat step (2) till all vertices from the spanning tree become included in the
# t-cherry tree.

g_CL <- g_final
#plot(g_CL, layout = layout_as_tree)


#FIXME:
# Calc MI(i, j, k) and start in that order.

# Calculate three variable mutual information

# A2.1)
# The first cherry of the t-cherry tree let be defined by any three vertices of the
# spanning tree which are connected by two edges.
# Tree -- so start at random node, take random neibour and again random neigbour (but not the first)

g_tCT <- g_CL

set.seed(1000)
first_cherry_i <- sample(names(V(g_CL)), 1L)
first_cherry_j <- sample(names(neighbors(g_CL, first_cherry_i)), 1L)
first_cherry_k <- sample(setdiff(names(neighbors(g_CL, first_cherry_j)), first_cherry_i), 1L) # except i again


# first cherry already in tree!
first_cherry <- c(first_cherry_i, first_cherry_j, first_cherry_k)
first_cherry 

cherries <- list(first_cherry)
# vertices_left <- setdiff(names(V(g_CL)), first_cherry)
# vertices_left
get_adjencent_vertices <- function(graph, cherries) {
  adjencent_vertices <- unique(unlist(lapply(cherries, function(vs) {
    lapply(vs, function(v) {
      names(neighbors(graph, v))
    })
  })))
  adjencent_vertices
}
adjencent_vertices <- get_adjencent_vertices(g_tCT, cherries)
#adjencent_vertices
adjencent_vertices_not_in_cherry_already <- setdiff(adjencent_vertices, unlist(cherries))
adjencent_vertices_not_in_cherry_already

# A2.2. We add a new cherry to the t-cherry tree by taking a new vertex of the spanning
# tree adjacent to the so far constructed t-cherry tree.
#    Keep track of neighbours of existing cherries (not already in a cherry).
#       -> Pick one of these
#    t cherry (not just cherry), so take random edge (two nodes that are connected) and the nodes there
plot(g_tCT, layout = layout_as_tree)

while (length(adjencent_vertices_not_in_cherry_already) >= 1L) {
  # Pick an adjecent vertex by
  new_cherry_middle <- sample(adjencent_vertices_not_in_cherry_already, 1L)
  new_cherry_middle
  
  # Remove selected new middle to get edges without the selected middle
  g_tCT_minus_new_middle <- delete_vertices(g_tCT, new_cherry_middle)
  
  # Pick two vertices that are connected -- same as picking a random edge; FIXME: random/MI
  random_edge <- sample(E(g_tCT_minus_new_middle), 1)
  random_edge_ends <- ends(g_tCT_minus_new_middle, random_edge)
  
  new_cherry_left <- random_edge_ends[1L, 1L]
  new_cherry_right <- random_edge_ends[1L, 2L]
  
  new_cherry_left
  new_cherry_right
  
  new_cherry <- c(new_cherry_left, new_cherry_middle, new_cherry_right)
  new_cherry
  
  ##########################
  
  # Only add edges if not existing:
  # The vertex chosen is already connected, so one edge must already exist.
  # Only add if not already present.
  
  if (!are.connected(g_tCT, new_cherry_left, new_cherry_middle)) {
    g_tCT <- g_tCT + 
      edge(new_cherry_left, new_cherry_middle)
  }
  
  if (!are.connected(g_tCT, new_cherry_middle, new_cherry_right)) {
    g_tCT <- g_tCT + 
      edge(new_cherry_middle, new_cherry_right)
  }

  plot(g_tCT)
  
  cherries[[length(cherries) + 1]] <- new_cherry
  adjencent_vertices <- get_adjencent_vertices(g_tCT, cherries)
  adjencent_vertices_not_in_cherry_already <- setdiff(adjencent_vertices, unlist(cherries))
  
  #break
}

plot(g_CL)
#plot(g_tCT, layout = layout_as_tree)

# FIXME:
# All already present cherries from Chow-Liu not registered... Should they be in 
# cherries as well?

# Remark: 3.2: n vertices, n-2 cherries:
length(cherries)
ncol(db)
stopifnot(length(cherries))

library(gRbase)
g <- igraph::igraph.to.graphNEL(g_tCT)
g
g_rip <- rip(g)
g_rip
#plot(g_rip)

sapply(g_rip$cliques, length)
stopifnot(all(sapply(g_rip$cliques, length) == 3L))

sapply(g_rip$separators, length)
stopifnot(all(sapply(g_rip$separators, length) == 2L))

```



```{r}
# A2.1: 
# three vertices of the spanning tree which are connected by two edges.
```

